Neural Co-Processors for Restoring Brain Function:\\Results from a Cortical Model of Grasping

### 2022-10-22 paper/article.tex
JNE paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{graphicx,epstopdf}
\usepackage{caption}
\usepackage{subcaption}
\graphicspath{ {./figs/} }
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\begin{document}

\title[Modeling a Neural Co-Processor]
{Towards a Neural Co-Processor Which Restores Dexterity After Stroke: Modeling a Proof-of-Concept}

\author{Matthew J Bryan$^{1}$, Linxing Preston Jiang$^{1}$, Rajesh P N Rao$^{1}$}

\address{$^{1}$ Neural Systems Laboratory, Department of Computer
Science and Engineering, University of Washington, Box 352350,
Seattle, WA 98105, USA}

\ead{matthew.bryan@dell.com}
\vspace{10pt}
\begin{indented}
\item[]September 2021
\end{indented}

\begin{abstract}
\textit{Objective} Brain co-processors \cite{rao.coproc} are devices which use artificial
intelligence (AI) for closed-loop neurostimulation, to shape neural activity and to bridge
injured neural circuits for targeted repair and rehabilitation. The co-processor framework
offers a flexible approach to learning closed-loop stimulation policies that optimize for
(a) specific regimes of neural activity, or (b) external task performance.
For example, one may seek to learn to stimulate the motor cortex of a stroke patient,
conditioning the stimulation on upstream visual information, aiding the patient's
attempt to grasp an object. Through the use of artificial neural
networks (ANNs) and deep learning, the co-processor co-adapts with the
neural circuit, allowing it to seek optimal stimulation policies, and adapt them
as the neural circuit changes. The results presented here demonstrate a
neural co-processor for the first time, through the use of a simulation. We
explore some of the core algorithms that may allow co-processors to successfully
learn, and to adapt to non-stationarity in both the brain and sensors.
\textit{Approach} We provide the first proof-of-concept of a neural co-processor that
leverages deep learning, through the use of a simulated neural circuit.
That circuit performs a reach-to-grasp task, based on visual input, and is designed to
closely resemble a similar circuit in a primate brain \cite{michaels.mrnn}. We simulate
a variety of lesions by altering the model, and then demonstrate our co-processor's ability
to restore lost function through ``stimulation'' of that model. We further test the
ability of our co-processor to adapt its stimulation as the simulated brain undergoes changes.
\textit{Main results} Our simulated co-processor successfully co-adapts with the neural
circuit to accomplish the external reaching task. The co-processor framework
demonstrated here adapts to a variety of lesion types, and to ongoing changes in the
simulated brain.
\textit{Significance} The proof-of-concept here outlines a co-processor model, as well
as our approach to training it, leading to insights on how such a model may be
developed for \textit{in vivo} use. We believe this co-processor design has potential to
learn complex stimulation policies, leading to application in treating a variety of
diseases.
\end{abstract}

\vspace{2pc}
\noindent{\it Keywords}: brain-computer interface, neural co-processor, ai, machine learning, stimulation
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction}
Aided in part by application of advanced AI techniques, brain-computer interfaces (BCIs) have made
advancements over the last several decads, allowing for decoded brain signals to be used for
control of a wide variety of virtual and physical prostheses \cite{rao.bcibook, wolpaw.bcibook,
moritz.neuro, lebedev.bmi}. Separately: advances in stimulation techniques and modeling have allowed
us to probe neural circuit dynamics (e.g. \cite{walker.inception}) and learn to better drive neural
circuits towards target dynamics, by encoding and delivering information through
stimulation \cite{niparko.cochlear, weiland.retinal, tomlinson.propr, tabot.tact, tyler.tact,
dadarlat.tact, sharlene.tact, cronin.tact}.
Recently, there has been increasing interest in building on these advances to combine decoding
and encoding in a single system, for closed-loop stimulation of a neural circuit. Bi-directional
BCIs (BBCIs) allow stimulation to be conditioned by decoded brain activity as well as external
sensor (e.g. camera) data, which can allow for the application of real-time, fine-grained control of
neural circuits and prosthetic devices, e.g. Nicolelis et al. \cite{nicolelis.bmbi}. These may lead,
for example, to neuro-prostheses that are capable of restoring movement which was lost due to
traumatic brain injury (TBI), to a degree not previously possible.

Motivated by that progress, we demonstrate here a flexible framework for combining encoding
and decoding, which we term ``neural co-processors'' \cite{rao.coproc}. Neural co-processors leverage
AI and deep learning to identify optimal, closed-loop stimulation patterns. The approach is flexible
enough to optimize not only for particular neural activities, but also for tasks external to the
subject. For example, they may be able to aid a stroke patient by finding a stimulation pattern of
the motor cortex which helps restore lost limb function. Likewise, the framework generalizes enough
to condition stimulation on both brain activity, and external sensors, e.g. cameras or light detection
and ranging (LIDAR) sensors, in order to incorporate feedback for realtime control.

Additionally, the co-processor framework allows a neuro-prosthesis to actively adapt to
a neural circuit as it changes with time. This framework is capable of co-adapting with
the circuit, i.e. brain, by updating its stimulation regime, while at the same time the
brain is updating its response to the stimulation, and changing due to natural plasticity
and aging. This allows the co-processor to continually optimize for the intended cost
function, despite the signficant non-stationarity of the target circuit.

Here we provide a proof-of-concept in simulation for a co-processor that restores
movement to a limb, after a subject has suffered a stroke affecting their ability to
use that limb. It combines:
\begin{itemize}
	\item A stimulation model, which models the relationship between decoded brain activity,
	      stimulation, and task performance.
	\item An AI agent which determines the stimulation to apply in a closed-loop fashion, in real time.
\end{itemize}

\section{Background}
Significant advances have been made in modeling the effects of electrical stimulation
of the brain, some of which can be leveraged for our co-processor design, as we outline
below. Researchers have explored how information can be biomimetically or
artificially encoded and delivered via stimulation to neuronal networks in the brain and
other regions of the nervous system for auditory \cite{niparko.cochlear}, visual \cite{weiland.retinal},
proprioceptive \cite{tomlinson.propr}, and tactile
\cite{tabot.tact, tyler.tact, dadarlat.tact, sharlene.tact, cronin.tact} perception.
Advancements have also been made in modeling the effects of stimulation over large scale, multi-region
networks, and across time \cite{shanechi.stimmodel}. Some have additionally designed models which
can adapt to ongoing changes in the brain, including changes due to the stimulation itself
\cite{tafazoli.acls}. In our proof-of-concept outlined below, we will use a stimulation
model, not unlike those cited here, which seeks to account for both network dynamics
and non-stationarity. In addition to training the model to have a strong ability to predict
the effect of stimulation, we additionally train it to be useful for then learning an
optimal stimulation policy, which is a property somewhat distinct from predictive
power alone.

Advances have also been made in both open- and closed-loop stimulation for
treating a variety of disorders. Open loop stimulation has been effective in
treating Parkinson's Disease (PD) \cite{benabid.parkinsons}, as well as various
psychiatric disorders \cite{holtzheimer.psy, kisely.psy, fraint.psy}.
More directly related to this paper, we see in Khanna et al. \cite{khanna.openloop},
the use of open loop stimulation in restoring dexterity after a lesion
occurs affecting a primate's motor cortex. The authors demonstrate that
the use of low-frequency alternating current, applied epidurally,
can improve grasp performance.

While open loop stimulation techniques have yielded clinically useful results,
their results in many domains have been mixed, such as use in visual
prostheses \cite{bosking.visual}, and use in invoking somatosensory feedback
\cite{cronin.tact}. Likely this is due to the stimulation not being conditioned
on the ongoing dynamics of the circuit being stimulated. Moment-to-moment and
throughout the day, the circuit will respond differently to the same stimulus,
as a result of differing inputs and ongoing activity. Stimulation
therefore needs to be proactively adapted in response. This need is even
greater over longer time scales as the effects of plasticity and ageing change
the connectivity of the brain. Closed-loop stimulation may also provide means
to better regulate the energy use of an implanted stimulator, allowing it to
intelligently regulate when to apply stimulation, in order to preserve implant
battery life. Closed-loop stimulation also offers an opportunity minimize
side-effects of stimulation, through real time regulation
of the stimulation parameters, such as in the use of deep brain stimulation (DBS) in
PD patients \cite{little.park}.

Closed-loop stimulation conditions stimulation on observations of brain activity,
allowing it to shape the neural activity more precisely, and in response to changes
in a patient's clinical condition. This may also allow an implant to adapt
in real time to changes in the target circuit. It has been used to aid in learning
new memories after some impairment \cite{berger.closedloop, kahana.biomarker},
to replay visually-invoked activations \cite{tafazoli.acls}, and for optogenetic
control of a thalamocortical circuit \cite{bolus.opto}, among others.

Something that remains unclear is how to leverage closed-loop control for real-time
co-adaption with the brain to accomplish an external task. ``Co-adaption'' here refers
to the ability of a neuro-prosthesis to adapt its stimulation regime to the ongoing
changes in the circuit it is stimulating, and to adapt with that circuit to accomplish
the external task, such as grasping. The neural co-processor we present here provides one
potential model for accomplishing that. Through the use of deep learning, the
co-processor model we present co-adapts an AI agent, which governs the
stimulation, with both a stimulation model, and the neural circuit being stimulated.

For a neurologically complex task such as grasping, we cannot identify
\textit{a priori} a real time controller of the neural circuit. That is
due in large part to the variability of circuits from subject-to-subject,
as well as variations in the placement of sensors and stimulators in the
brain. The only plausible path to such a real time controller is to parameterize
it in a subject- and time-specific way. Our model seeks to accomplish that using
deep learning.

Before attempting \textit{in vivo} experiments using such a model, we first demonstrate
here a number of crucial design elements of it, through the use of a simulated
grasping circuit, presented previously by Michaels et al. \cite{michaels.mrnn}. We
explore the properties of the co-processor design that are needed to
successfully adapt to the long-running dynamics of a stimulated neural circuit,
as well as to adapt to that circuit's ongoing connectivity changes.
We additionally present a training method for our co-processor, allowing it to learn
an optimal stimulation pattern that drives improvements in external task performance,
while also adapting to the non-stationarity of the brain.

\section{Method}

\begin{figure}
\includegraphics[width=\textwidth]{weill_arch.png}
\caption{Using a co-processor to drive external task performance after a traumatic brain injury}
\centering
\label{fig:weill}
\end{figure}

\subsection{Architecture Overview}
First, we present the architecture of our co-processor design. This design aims to solve two
fundamental challenges in using neural stimulation to improve external task performance.
First, neural networks exhibit long-running and nonlinear dynamics, necessitating
the need for a stimulation agent to account for far-distant effects of the stimulation it
applies. Second, with neurologically complex tasks, such as grasping, the mapping between
neural activity as-measured and the external task cannot be determined \textit{a priori}.
As a result, the co-processor must somehow learn what stimulation is appropriate for
aiding the user in the external task they are attempting to perform.

Our co-processor attempts to solve these with a pair of artificial neural networks:
\begin{itemize}
	\item A stimulation and neural dynamics model, known as an ``Emulator Network'' (EN). It models the relationship
	      between the stimulation, neural dynamics, and external task. Its purpose is for training the second network.
	\item A stimulation agent, known as the ``co-processor network'' (CPN), which maps neural activity, and possibly
	      data from external sensors, to stimulation parameters.
\end{itemize}

We co-train the EN and CPN, with the goal of training a CPN whose output stimulation parameters
improve task performance. By continually training both, they adapt to the brain as it changes,
effectively allowing for brain-stimulator co-adaptation. The EN is a tool for training the CPN,
giving us a way to backpropagate task error to the CPN. It outputs task-relevant
metrics - a prediction of muscle velocities in our case - given measurements of neural
activity, and the stimulation parameters output from the CPN. If the EN is trained in a
particular way, and to a sufficient level of precision, it can be used as a function
approximator relating stimulation and a task, and do so in a way that allows us to train
the CPN with it. When training the CPN, we in-effect treat the EN's output as the true
task performance, or a related metric, and then backpropagate the loss defined in terms
of that metric in order to train the CPN. See Fig. \ref{fig:weill}. We will illustrate the details
of the training algorithm below.

In our present demonstration, the EN is constituted as a single layer, fully connected,
long short-term memory (LSTM) recurrent neural network (RNN), with hyperbolic tangent ($tanh$)
activations, and a linear readout. The general notion of a co-processor does
not require this precise architecture, but for our example, we found that the LSTM
approach allows the network to continuously adapt to long-running dependencies in
the simulated neural dynamics, far better than a vanilla RNN. The CPN is constituted
as an almost identical network, though with a different dimensionality, as we explain
below. There is no strict requirement for the EN and CPN to be so similar, but we
found this simple architecture to work well in our example.

Note that the EN effectively constitutes a stimulation model. However, in the case that
a co-processor attempts to optimize for task performance, as opposed to neural
activity directly, the stimulation model predicts the effect of stimulation and neural
dynamics on task performance, rather than modeling and predicting the neural response
to stimulation alone. This provides the key functionality needed to train the CPN.

We initially attempted an EN architecture using vanilla RNN cells, with a nonlinearity.
This approach resembles the common linear time-invariant state space model of
stimulation, as in e.g. Yang et al. \cite{shanechi.stimmodel}, though it additionally
incorporates a nonlinearity. That approach has the benefit of being simple, and familiar.
We found, however, that the linear approach, as well as the related vanilla RNN
approach with nonlinearities added, was not sufficiently expressive of long-term
dependencies, such that the co-processor was capable of learning well.

Separately, nonlinear LSTM models have also been applied to modeling natural
neural networks, such as use in predicting local field potentials
\cite{kim.lstm}, and for stimulation modeling \cite{guclu.lstm}.
LSTM cells are designed to better capture the relevant aspects of
long-running network dynamics. We found that LSTM cells were crucial for
enabling our co-processor to learn the long-running dynamics of the network
it was stimulating, and we use them as a result.

\subsection{Simulation Overview}
We demonstrate our approach here using a simulated grasping circuit.
A detailed simulation such as this allows us to explore some of the critical
architectural details and training algorithms that will allow a co-processor
to succeed.  By first doing such exploration in simulation, we are able to
rapidly and cheaply iterate on our design, prior to any \textit{in vivo}
experiments. For the simulation to be admissible, we need to ensure it has some
properties that allow it to strongly indicate if our design is improving in
a direction that will later allow for real deployments. Otherwise, our
design may be adapting to the pecularities of the simulation, without
becoming more useful.

An example of such a simulation approach is Dura-Bernal et al. \cite{bernal.sim}. In this work, the
authors use a simulated spiking neural network to train a stimulation agent.
Their stimulation agent sought to restore the network's
control of a simulated arm, to reach a target, after a simulated lesion was applied,
much like the work here.  As the authors point out, we have a limited ability to
probe a neural circuit \textit{in vivo} in order to perform learning. As a result,
we first need to design our approach through the use of an admissible simulation.
The authors in this case simulated lesions by effectively removing parts of their
simulated network, or by cutting connections between parts of the network.
Here, we use an established design for an artificial neural circuit which
performs a grasping task. Its architecture and training methods were designed
specifically to result in naturalistic dynamics.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\usepackage{graphicx,epstopdf}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{lscape}
\usepackage{hyperref}
\graphicspath{ {./figs/} }
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\begin{document}

\title[Neural Co-Processors]
{Neural Co-Processors for Restoring Brain Function:\\Results from a Cortical Model of Grasping}

% Authorship temporarily removed for double-anonymous review
\author{Matthew J Bryan$^{1}$, Linxing Preston Jiang$^{1}$, Rajesh P N Rao$^{1}$}

\address{$^{1}$ Neural Systems Laboratory, Paul G. Allen School of Computer
Science \& Engineering, University of Washington, Box 352350,
Seattle, WA 98105, USA}

\ead{\{mmattb,prestonj,rao\}@cs.washington.edu}
\vspace{10pt}
\begin{indented}
\item[]November 2022
\end{indented}

\begin{abstract}
\textit{Objective} A major challenge in closed-loop brain-computer interfaces (BCIs)
is finding optimal stimulation patterns as a function of ongoing neural activity for
different subjects and for different objectives. Traditional approaches, such as
those currently used for deep brain stimulation (DBS), have largely followed a manual
trial-and-error strategy to search for effective open-loop stimulation parameters,
a strategy that is inefficient and does not generalize to closed-loop
activity-dependent stimulation.
\textit{Approach} To achieve goal-directed closed-loop neurostimulation, we propose
the use of brain co-processors, devices which exploit artificial intelligence (AI)
to shape neural activity and bridge injured neural circuits for targeted repair and
rehabilitation.  Here we investigate a specific type of co-processor called a ``neural
co-processor'' which uses artificial neural networks (ANNs) and deep learning to learn optimal
closed-loop stimulation policies. The co-processor adapts the stimulation policy as
the biological circuit itself adapts to the stimulation, achieving a form of brain-device
co-adaptation. We tested the neural co-processor's ability to restore function after
stroke by simulating a variety of lesions in a previously published cortical model
of grasping.  We additionally tested the ability of our co-processor
to adapt its stimulation as the simulated cortical network and simulated sensors
underwent changes.
\textit{Main results} Our results show that a neural co-processor can restore
reaching and grasping function after a simulated stroke in a cortical model, achieving
recovery towards healthy function in the range 75-90\%. Our co-processor successfully
co-adapted to accomplish the reach-and-grasp task after a variety of lesions to the
simulated cortical circuit.
\textit{Significance} Our results provide the first proof-of-concept demonstration,
using computer simulations, of a neural co-processor for activity-dependent closed-loop
neurosimulation for optimizing a rehabilitation goal after injury. Our results provide
insights on how such co-processors may eventually be developed for \textit{in vivo} use
to learn complex adaptive stimulation policies for a variety of neural rehabilitation
and neuroprosthetic applications. 
\end{abstract}

\vspace{2pc}
\noindent{\it Keywords}: brain-computer interface, neural co-processor, AI, machine learning, deep learning, neural networks, stimulation, computational models
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction}
Brain-computer interfaces (BCIs) have made significant advances over the last several decades, leading
to the control of a wide variety of virtual and physical prostheses through neural signal decoding
\cite{rao.bcibook, wolpaw.bcibook, moritz.neuro, lebedev.bmi}. Separately, advances in stimulation techniques and modeling have allowed
us to probe neural circuit dynamics (e.g. \cite{walker.inception}) and learn to better drive neural
circuits towards desired target dynamics by encoding and delivering information through
stimulation \cite{niparko.cochlear, weiland.retinal, tomlinson.propr, tabot.tact, tyler.tact,
dadarlat.tact, sharlene.tact, cronin.tact}. Bi-directional BCIs (BBCIs) allow stimulation to be
conditioned on decoded brain activity and encoded sensor data for applications such as real-time,
fine-grained control of neural circuits and prosthetic devices (e.g., \cite{nicolelis.bmbi}).

Motivated by these advances, we investigate here a flexible framework for combining encoding
and decoding using ``neural co-processors'' \cite{rao.coproc}, a type of brain co-processor \cite{rao.braincoproc}. Neural co-processors leverage artificial neural networks (ANNs) and deep
learning to compute optimal closed-loop stimulation patterns. The approach can be used to not only drive neural activity
toward desired activity regimes, but also to achieve task goals external to the subject, such as finding closed-loop
stimulation patterns for motor cortical neurons for restoring the ability to reach and grasp an object. Likewise, the
framework generalizes to stimulation based on both brain activity and external sensor measurements, e.g., from cameras or
light detection and ranging (LIDAR) sensors, in order to restore perception (e.g., cortical visual prosthesis) or
incorporate feedback for real-time prosthetic control (see \cite{rao.braincoproc} for details).

The co-processor framework also allows co-adaptation with biological circuits in the brain by updating its stimulation policy, while the brain updates its own response to the stimulation via adaptation and neural plasticity, or updates due to other reasons. This allows the co-processor to continually optimize its outputs for the desired optimization function in the presence of significant non-stationarity in the brain.

Here we use computer simulations to demonstrate a neural co-processor that restores
movement in a computational model of cortical networks controlling a limb, after a simulated stroke affects the ability to
use that limb. Our demonstration combines:
\begin{itemize}
	\item An emulation model based on ANNs, which models the relationship between 
	      stimulation, decoded brain activity, and task performance.
	\item An AI agent based on ANNs which determines the best stimulation to apply in a closed-loop fashion in real time.
\end{itemize}

\section{Background}
Significant advances have been made in understanding and modeling the effects of electrical stimulation
on the brain. Researchers have explored how information can be biomimetically or
artificially encoded and delivered via stimulation to neuronal networks in the brain and
other regions of the nervous system for auditory \cite{niparko.cochlear}, visual \cite{weiland.retinal},
proprioceptive \cite{tomlinson.propr}, and tactile
\cite{tabot.tact, tyler.tact, dadarlat.tact, sharlene.tact, cronin.tact} perception.
Advances have also been made in modeling the effects of stimulation over large scale, multi-region
networks, and across time \cite{shanechi.stimmodel}. Some models can additionally adapt to ongoing
changes in the brain, including changes due to the stimulation itself
\cite{tafazoli.acls}. For our simulations described below, we use a stimulation
model, not unlike those cited above, which seeks to account for both network dynamics
and non-stationarity. In addition to training the model to have a strong ability to predict
the effect of stimulation, we additionally adapt it to be useful for learning an
optimal stimulation policy, a property distinct from predictive
power alone.

Researchers have also explored both open- and closed-loop stimulation protocols for
treating a variety of disorders. Open loop stimulation has been effective in
treating Parkinson's Disease (PD) \cite{benabid.parkinsons}, as well as various
psychiatric disorders \cite{holtzheimer.psy, kisely.psy, fraint.psy}.
In research more directly related to our work, Khanna et al. \cite{khanna.openloop}, investigated 
the use of open loop stimulation in restoring dexterity after a lesion in a nonhuman primate's (NHP's) motor cortex.
The authors demonstrate that the use of low-frequency alternating current, applied epidurally,
can improve grasp performance.

While open loop stimulation techniques have yielded clinically useful results, results in other domains have been mixed, such as in visual
prostheses \cite{bosking.visual}, and in invoking somatosensory feedback
\cite{cronin.tact}. We believe this is due to the stimulation not being conditioned
on the ongoing dynamics of the neural circuit being stimulated. From moment to moment and
throughout the day, a neuronal circuit in the brain can be expected to respond differently even when the same stimulation parameters are used,
due to the multitude of different external and internal inputs influencing the circuit's ongoing activity. Stimulation
therefore needs to be proactively adapted in response. This need is even
greater over longer time scales as the effects of plasticity, changes in clinical conditions, and ageing change
the dynamics and connectivity of the brain. Closed-loop stimulation may also provide means
to better regulate the energy use of an implanted stimulator, allowing it to
intelligently regulate when to apply stimulation, in order to preserve implant
battery life. Another benefit is that closed-loop stimulation offers an opportunity to minimize the 
side-effects of stimulation, through real time regulation
of the stimulation parameters, such as in the use of deep brain stimulation (DBS) in
PD patients \cite{little.park}. In recent years, closed-loop stimulation has been used to aid in learning
new memories after some impairment \cite{berger.closedloop, kahana.biomarker},
to replay visually-invoked activations \cite{tafazoli.acls}, and for optogenetic
control of a thalamocortical circuit \cite{bolus.opto}, among others.

A major open question is: how does one leverage closed-loop control for real-time
co-adaptation with the brain to accomplish an external task? ``Co-adaptation'' here refers
to the ability of a BCI to adapt its stimulation regime to the ongoing
changes in the circuit it is stimulating, and to adapt with that circuit to accomplish
the external task (e.g., grasping). The neural co-processor we present here provides one
potential model for accomplishing that. Through the use of deep learning, a neural 
co-processor model co-adapts its AI, which governs the
stimulation, with the neural circuit being stimulated.

For a neurologically complex task such as grasping, it is not possible to design
\textit{a priori} a fixed real-time controller for the (potentially impaired) neural circuits involved in the task. That is
due in large part to the variability of circuits from subject-to-subject,
as well as variations in the placement of sensors and stimulators in the
brain. The only plausible path to implementing a real-time controller is therefore to allow the device to adapt to the subject,
long-term changes in their brain activities, and variability in the sensors, stimulators and hardware. Our proposed neural
co-processors seek to accomplish such adaptation through ANNs and deep learning.

To gain insights into neural co-processors before testing them in \textit{in vivo} experiments, we investigated a number of crucial design elements through the use of a neural network model of the cortical areas involved in 
grasping, presented previously by Michaels et al. \cite{michaels.mrnn}. We
explored what properties of the co-processor allow
successful adaptation to the short-term dynamics of the cortical model as it is being stimulated,
as well as adaptation to longer-term connectivity changes in the cortical model.
We present a training method for neural co-processors for learning optimal stimulation patterns that drive improvements in external task performance,
while also adapting to the non-stationarity of the brain.

\section{Methods}

\begin{figure}
    \centering
    \begin{subfigure}[c]{0.75\textwidth}
		\centering
		\includegraphics[width=\textwidth]{weill_arch.png}
		\caption{}
	\end{subfigure}
	\hfill
    \begin{subfigure}[c]{0.75\textwidth}
		\centering
		\includegraphics[width=\textwidth]{cpn_michaels_arch_labeled.png}
		\caption{}
	\end{subfigure}
	\hfill
    \caption{{\bf Neural co-processor for restoring function after a brain injury}.
    An artificial neural network called the “Co-Processor Network” (CPN) is used to
    map input neural activity patterns in an area A to output stimulation patterns
    in same or other areas B in order to achieve a neural or behavioral goal using
    another ANN, an “Emulator Network” (EN) - see text for details. (a) The example
    here shows the CPN creating a new information processing pathway between prefrontal
    cortex and motor cortex, bypassing an intermediate area affected by brain injury
    (e.g., stroke). Adapted from \cite{rao.coproc}, (b) Our current study involves a
    simulated cortical grasping circuit. The CPN and EN both receive simulated brain
    recordings, which are created according to a recording model (see Methods).
    The CPN outputs stimulation parameters, which are applied to the simulated
    grasping circuit, according to a stimulation model. The EN models the relationship between the
    stimulation, brain recordings, and external task.}
    \label{fig:arch}
\end{figure}

\subsection{Architecture Overview}
First, we present the architecture of our neural co-processor design. This design aims to solve two
fundamental challenges in using neural stimulation to improve external task performance.
First, to restore function for complex tasks, such as grasping, the mapping between
neural activity, such as the activity encoding the intention to grasp, and the stimulation patterns
to be applied to a downstream motor area to enable grasping cannot be pre-determined.
As a result, the co-processor must learn what stimulation pattern is appropriate for achieving the
external task. Unfortunately, in most cases, we do not know the optimal stimulation patterns to train
the co-processor. This is because stimulation shapes the nonlinear dynamics of circuits in the brain
in complex ways, leading to complex effects in external behavior. It is therefore not obvious what
stimulation patterns will produce a desired behavior.

A neural co-processor attempts to solve these problems with a pair of artificial neural networks (ANNs)
(Fig.~\ref{fig:arch}):
\begin{itemize}
    \item a ``Co-Processor Network'' (CPN), which maps neural activity, and possibly
	      data from external sensors, to appropriate stimulation parameters.
    \item an ``Emulator Network'' (EN) which models the effect of stimulation on neural dynamics and
          behavior for the external task.
\end{itemize}

The CPN can be trained using the backpropagation algorithm, the workhorse of deep learning for training ANNs.
However, backpropagation requires the error between the output of the CPN and a desired output, and as
discussed above, we do not have the desired output stimulation pattern. We do however know what the desired
output behavior in a task should be, e.g., a particular kind of grasp for a particular object or a particular
type of neural activity in a brain area corresponding to healthy activity. We can therefore compute the error
between this desired behavior and the actual behavior caused by the CPN due to stimulation. How do we
backpropagate this error, which is external to the CPN, to update the parameters of the CPN?

We use a trained EN to backpropagate the external task error to the CPN. We train the EN to predict
task-relevant parameters - a prediction of muscle velocities in our case - given the stimulation
parameters output by the CPN and any measured neural activity. If the EN is trained to a sufficiently
high level of precision, it can be used as a function approximator mapping stimulation to task output.
When training the CPN, we treat the EN's output as the actual task output (e.g., actual grasp
parameters).  We backpropagate through the EN (without changing its weights) the error between
the EN's output and the desired task output and use this backpropagated error to train the
CPN (see Fig. \ref{fig:arch}).

In our experiments, the EN was a single layer fully connected long short-term memory (LSTM) recurrent
neural network (RNN), with hyperbolic tangent ($tanh$) activations, and a linear readout. It has 87 LSTM neurons,
which was chosen somewhat arbitrarily as a function of the input and output vector sizes. We found varying this neuron
count did not drastically change results. Although other architectures could also be used, we found that this
LSTM architecture allows the EN to continuously adapt to long-running dependencies in the simulated neural dynamics, 
ar better than a vanilla RNN. The CPN had an almost identical architecture, but with 61 LSTM neurons. As with the CPN,
this neuron count was chosen as a function of the input and output vector sizes, but varying it upward had little effect on results.
There is no requirement for the EN and CPN to have similar network architectures, but we found that these choices worked
well in our experiments.

Note that the EN is more general than traditional models of neurostimulation which attempt to predict the effects of
stimulation on \em{neural activity}. The EN in a neural co-processor predicts the effects of stimulation (taking into
account ongoing neural dynamics) on \em{task performance}. This provides the key functionality needed to train the
CPN. In the special case where the task involves driving neural circuits in the brain to desired neural activities,
the EN reduces to more traditional models of stimulation.

For comparison, consider an EN architecture based on a traditional RNN with a nonlinearity. This is a nonlinear version
of the common linear time-invariant state space model of stimulation (see, e.g.,  \cite{shanechi.stimmodel}). 
We found, however, that compared to LSTM-based EN, the linear model and the traditional nonlinear RNN are both not
sufficiently powerful to capture the long-term dependencies in stimulation effects needed for the CPN to learn well.
Our use of an LSTM-based EN builds on previous work modeling biological neural networks for predicting local field potentials
\cite{kim.lstm}, and modeling stimulation \cite{guclu.lstm}.

\subsection{Simulation Overview}
To test the feasibility of the neural co-processor approach, we used a previously published model of a
grasping circuit in a nonhuman primate (NHP) brain. Using such a model allows us to explore some of the critical
architectural choices and training algorithms, enabling us to rapidly and cheaply iterate on our
design, prior to any \textit{in vivo} experiments.

In the present study, we use the grasping network model of Michaels et al. \cite{michaels.mrnn}
(Fig.~\ref{fig:michaels}a). The network, representing multiple interconnected cortical areas, was trained
to mimic the grasping circuits of NHP subjects engaged in a delayed reach-to-grasp task.
The model's design draws on a body of literature focused on architectures and training methods for RNNs which
seek to create artificial neural networks with activation dynamics similar to biological circuits,
including circuits for delayed grasping tasks \cite{susillo.mrnn}. The model consists of three
``modular'' vanilla RNNs (mRNNs), representing the cortical areas AIP, F5 and M1 respectively, and a linear
readout layer (Fig.~\ref{fig:michaels}a). Each ``module'' consists of 100 vanilla
RNN neurons, with a nonlinearity applied on the outputs. The modules are internally fully connected,
and are connected to each other sparsely (10\% connectivity in our case). The inputs to the network are visual features
representing the object to be grasped, as well as a hold signal. The visual features
intend to capture the features represented in the subject's visual cortex. They were extracted using
VGGNet \cite{simonyan.vgg} from 3D renderings of the same objects which the subjects grasped. The hold signal is
a boolean which encodes the point in the experiment when the subject began to reach. The outputs of the
network are muscle length velocities for the shoulder, arm, and hand of the subject. The actual
velocities were captured with a motion capture glove, and the Michaels et al. network model 
was trained to recapitulate those grasping motions.\footnote{Data and trained models from this work
were supplied to us by the lead author of \cite{michaels.mrnn}. We re-implemented their model in PyTorch,
and used their trained parameters for an arbitrarily chosen subject.}
